---
title: "Introduction to linkinfer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to linkinfer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

`linkinfer` provides tools for researchers working with linked administrative
or census data. When records are linked across datasets, two types of errors
can affect inference:

1. **Missed matches** – records that should have linked but didn't, creating
   non‑random sample selection.
2. **False matches** – records incorrectly linked to the wrong individual,
   introducing measurement error.

This package helps diagnose and correct for these errors following the
framework of Breen and Joo (2026), "Population Research With Linked Data:
Methods for Inference."

## Setup

```{r setup, include=FALSE}
# Reload the development version of linkinfer, ensuring no older version is attached
if ("linkinfer" %in% loadedNamespaces()) {
  detach("package:linkinfer", unload=TRUE, character.only=TRUE)
}
devtools::load_all('..')
```


```{r seed, echo=TRUE}
# Ensure reproducibility for the vignette
set.seed(1234)
```

## Example data

The package includes simulated data representing a census‑to‑linked records.
`population` is the full simulated census population, and `linked_sample`
is the subset that was successfully linked.

```{r data}
# Simulate data (the function adds stochastic noise to `linked_prob`)
sim <- linked_data(n = 10000, p = 0.2, v1 = 0.5, v2 = 0.8)
population <- sim$population
linked_sample <- sim$linked

cat("Population size:", nrow(population), "\n")
cat("Linked sample size:", nrow(linked_sample), "\n")
cat("Match rate:", round(nrow(linked_sample) / nrow(population), 3), "\n")
```

## Step 1: Diagnose representativeness

```{r compare}
comparison <- compare_linked(
  population = population,
  linked = linked_sample,
  vars = c("age", "sex", "race", "education", "urban", "name_commonness")
)
comparison
```

## Step 2: Adjust for missed matches

### Inverse‑probability weighting

```{r ipw}
ipw_result <- weight_ipw(
  population = population,
  linked_indicator = "linked",
  covariates = c("age", "sex", "race", "education", "urban", "name_commonness"),
  method = "logistic"
)
ipw_result
```

### Diagnostics (balance plot)

```{r ipw-diag}
diag <- weight_diagnostics(
  weights = ipw_result$weights,
  population = population,
  linked = linked_sample,
  covariates = ipw_result$covariates
)
diag
```

### Post‑stratification

```{r poststrat}
ps_result <- weight_poststratify(
  population = population,
  linked = linked_sample,
  strata_vars = c("sex", "race")
)
ps_result

ps_diag <- weight_diagnostics(
  weights = ps_result$weights,
  population = population,
  linked = linked_sample,
  covariates = c("sex", "race")
)
ps_diag
```

### Raking

```{r raking}
rake_result <- weight_rake(
  linked_data = linked_sample,
  population = population,
  vars = c("sex", "race", "education")
)
rake_result
```

## Step 3: Adjust for false matches using a validation variable

The function `adjust_estimate()` defaults to using ordinary least squares (`lm()`) for model‑based adjustments, but you can supply any fitting function that accepts a `formula`, a `data` frame, and a `weights` argument (e.g., `glm()`, `coxph()`, `lmer()`). Just provide the fitting function via the `model_fit` argument and a custom `extract_estimate` function to pull the scalar of interest from the fitted model.

```{r adjust-race-diff}
adjust_res <- adjust_estimate(
  weight_result = ipw_result,
  linked_data = linked_sample,
  population_data = population,
  outcome = "race_diff",
  validation_var = "middle_initial_match",
  n_boot = 200
)
adjust_res
```

### Regression‑based adjustment (OLS coefficient for Black vs White)

```{r adjust-regression, warning=FALSE, message=FALSE}
# Create a factor for race with White as reference
linked_sample$race_f <- factor(linked_sample$race, levels = c("White", "Black", "Other"))

model_form <- age_at_death ~ race_f

adjust_reg <- adjust_estimate(
  weight_result = ipw_result,
  linked_data = linked_sample,
  population_data = population,
  outcome = NULL,
  validation_var = "middle_initial_match",
  model_formula = model_form,
  extract_estimate = function(fit) coef(fit)["race_fBlack"],
  n_boot = 200
)
adjust_reg
```

## Summary

1. **Compare** linked vs. population characteristics with `compare_linked()`.
2. **Adjust** for missed matches using `weight_ipw()`, `weight_poststratify()`, or `weight_rake()`.
3. **Correct** for false matches using `adjust_estimate()` (validation‑variable based adjustment).

## References

Breen, C.F. and Joo, W. (2026). "Population Research With Linked Data:
Methods for Inference." Working paper.
