---
title: "Introduction to linkinfer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to linkinfer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

`linkinfer` provides tools for researchers working with linked administrative
or census data. When records are linked across datasets, two types of errors
can affect inference:

1. **Missed matches** – records that should have linked but didn't, creating
   non‑random sample selection.
2. **False matches** – records incorrectly linked to the wrong individual,
   introducing measurement error.

This package helps diagnose and correct for these errors following the
framework of Breen and Joo (2026), "Population Research With Linked Data:
Methods for Inference."

A useful way to think about the workflow is in three steps. Researchers first
diagnose whether the linked sample is representative of the source population
and, if not, apply weighting methods to address missed matches. They then
evaluate whether false matches are likely to bias the estimate of interest
and, when possible, use an auxiliary validation variable to estimate and correct
that bias. Finally, they report both the original and corrected estimates
so that the implications of linkage error are transparent.


## Setup

```{r setup, include=FALSE}
# Reload the development version of linkinfer, ensuring no older version is attached
if ("linkinfer" %in% loadedNamespaces()) {
  detach("package:linkinfer", unload=TRUE, character.only=TRUE)
}
devtools::load_all('..')
```


```{r seed, echo=TRUE}
# Ensure reproducibility for the vignette
set.seed(1234)
```

## Example data

The package includes simulated data representing a census‑to‑linked records.
`population` is the full simulated census population, and `linked_sample`
is the subset that was successfully linked. In this example, the linked sample
is affected by both missed and false matches.

The simulation also includes a validation variable, middle_initial_match,
which is observed only for a subset of linked records. This reflects a common
empirical setting. Researchers often do not observe ground truth for whether
each linked pair is correct, but they may observe an auxiliary variable that
provides information about match quality for part of the sample. Here, matching
middle initials serve as a proxy for higher-confidence links and are used to
estimate how false matches may be biasing the full-sample estimates.

```{r data}
# Simulate data (the function adds stochastic noise to `linked_prob`)
sim <- linked_data(n = 10000, p = 0.2, v1 = 0.5, v2 = 0.8)
population <- sim$population
linked_sample <- sim$linked

cat("Population size:", nrow(population), "\n")
cat("Linked sample size:", nrow(linked_sample), "\n")
cat("Match rate:", round(nrow(linked_sample) / nrow(population), 3), "\n")
```

## Step 1: Diagnose representativeness

The first step is to ask whether the linked sample resembles the population from
which it was drawn. If linkage success varies by age, sex, race, education, or other
characteristics, then inference based on the linked sample may be biased even before
considering false matches.

```{r compare}
comparison <- compare_linked(
  population = population,
  linked = linked_sample,
  vars = c("age", "sex", "race", "education", "urban", "name_commonness")
)
comparison
```

## Step 2: Adjust for missed matches

One way to correct for missed matches is to estimate each record’s probability of
appearing in the linked sample and then weight linked observations by the inverse of
that probability. This gives more influence to records from groups that were harder to link.

### Inverse‑probability weighting

```{r ipw}
ipw_result <- weight_ipw(
  population = population,
  linked_indicator = "linked",
  covariates = c("age", "sex", "race", "education", "urban", "name_commonness"),
  method = "logistic"
)
ipw_result
```

### Diagnostics (balance plot)

After estimating weights, it is useful to check whether the weighted linked sample is
more similar to the population.

```{r ipw-diag}
diag <- weight_diagnostics(
  weights = ipw_result$weights,
  population = population,
  linked = linked_sample,
  covariates = ipw_result$covariates
)
diag
```

### Post‑stratification

If population totals are known for discrete subgroups, post-stratification offers
a simpler correction.

```{r poststrat}
ps_result <- weight_poststratify(
  population = population,
  linked = linked_sample,
  strata_vars = c("sex", "race")
)
ps_result

ps_diag <- weight_diagnostics(
  weights = ps_result$weights,
  population = population,
  linked = linked_sample,
  covariates = c("sex", "race")
)
ps_diag
```

### Raking

When exact joint cells become sparse, raking provides a more flexible way to align
the linked sample to known marginal distributions.

```{r raking}
rake_result <- weight_rake(
  linked_data = linked_sample,
  population = population,
  vars = c("sex", "race", "education")
)
rake_result
```

## Step 3: Adjust for false matches using a validation variable

The function `adjust_estimate()` defaults to using ordinary least squares (`lm()`)
for model‑based adjustments, but you can supply any fitting function that accepts
a `formula`, a `data` frame, and a `weights` argument (e.g., `glm()`, `coxph()`,
`lmer()`). Just provide the fitting function via the `model_fit` argument and a custom
`extract_estimate` function to pull the scalar of interest from the fitted model.

```{r adjust-race-diff}
adjust_res <- adjust_estimate(
  weight_result = ipw_result,
  linked_data = linked_sample,
  population_data = population,
  outcome = "race_diff",
  validation_var = "middle_initial_match",
  n_boot = 200
)
adjust_res
```

### Regression‑based adjustment (OLS coefficient for Black vs White)

```{r adjust-regression, warning=FALSE, message=FALSE}
# Create a factor for race with White as reference
linked_sample$race_f <- factor(linked_sample$race, levels = c("White", "Black", "Other"))

model_form <- age_at_death ~ race_f

adjust_reg <- adjust_estimate(
  weight_result = ipw_result,
  linked_data = linked_sample,
  population_data = population,
  outcome = NULL,
  validation_var = "middle_initial_match",
  model_formula = model_form,
  extract_estimate = function(fit) coef(fit)["race_fBlack"],
  n_boot = 200
)
adjust_reg
```

## Summary

1. **Compare** linked vs. population characteristics with `compare_linked()`.
2. **Adjust** for missed matches using `weight_ipw()`, `weight_poststratify()`, or `weight_rake()`.
3. **Correct** for false matches using `adjust_estimate()` (validation‑variable based adjustment).

## References

Breen, C.F. and Joo, W. (2026). "Population Research With Linked Data:
Methods for Inference." Working paper.
