---
title: "Introduction to linkinfer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to linkinfer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

`linkinfer` provides tools for researchers working with linked administrative
or census data. When records are linked across datasets, two types of errors
can affect inference:

1. **Missed matches** — records that should have linked but didn't, creating
   non-random sample selection.
2. **False matches** — records incorrectly linked to the wrong individual,
   introducing measurement error.

This package helps diagnose and correct for these errors following the
framework of Breen and Joo (2026), "Population Research With Linked Data:
Methods for Inference."

## Setup

```{r setup}
library(linkinfer)
```

## Example data

The package includes simulated data representing a census-to-mortality linkage.
`census_population` is the full census (5,000 individuals), and
`linked_sample` is the subset successfully linked to mortality records.

```{r data}
data(census_population)
data(linked_sample)

cat("Population size:", nrow(census_population), "\n")
cat("Linked sample size:", nrow(linked_sample), "\n")
cat("Match rate:", round(nrow(linked_sample) / nrow(census_population), 3), "\n")
```

## Step 1: Diagnose representativeness

The first step is to compare the linked sample against the full population.
Are certain groups over- or under-represented?

```{r compare}
comparison <- compare_linked(
  population = census_population,
  linked = linked_sample,
  vars = c("age", "sex", "race", "education", "urban", "name_commonness")
)

comparison
```

Variables with large standardized differences (|std_diff| > 0.1) indicate
that the linked sample is not representative along that dimension. Here,
`name_commonness` stands out — individuals with common names are harder
to link uniquely, so they are underrepresented.

## Step 2: Adjust for missed matches

### Option A: Inverse probability weighting

We model the probability that each record was linked as a function of
observed characteristics, then weight linked records by the inverse of
that probability.

```{r ipw}
ipw_result <- weight_ipw(
  population = census_population,
  linked = linked_sample,
  covariates = c("age", "sex", "race", "education", "urban", "name_commonness"),
  method = "logistic"
)

ipw_result
```

Check whether the weights improve balance:

```{r ipw-diag}
diag <- ipw_diagnostics(ipw_result, census_population, linked_sample)
diag
```

If `ggplot2` is installed, you can produce a Love plot:

```{r love-plot, eval = requireNamespace("ggplot2", quietly = TRUE)}
plot(diag, type = "balance")
```

### Option B: Post-stratification

For a simpler approach, post-stratification reweights the linked sample so
that the distribution across cells matches the population (Eq. 1 in the paper:
$w_i = N_z / n_z$).

```{r poststrat}
ps_result <- weight_poststratify(
  population = census_population,
  linked = linked_sample,
  strata_vars = c("sex", "race")
)

ps_result
```

### Option C: Raking

When many adjustment variables are used, direct cross-classification leads
to sparse cells. Raking (iterative proportional fitting) adjusts to marginal
distributions instead. This uses the `autumn` package as a backend.

```{r raking, eval = requireNamespace("autumn", quietly = TRUE)}
# Compute population marginals
targets <- make_targets(
  data = census_population,
  vars = c("sex", "race", "education")
)

# Apply raking to the linked sample
rake_result <- weight_rake(
  linked_data = linked_sample,
  targets = targets
)

rake_result
```

## Step 3: Validate adjustments

A good validation variable is one that (a) is known at the population level
and (b) was **not** used to construct the weights. Here we use `urban`
residence as a validation variable, after fitting IPW weights that exclude
`urban` from the covariates:

```{r validate}
# Fit IPW *without* urban in the model
ipw_no_urban <- weight_ipw(
  population = census_population,
  linked = linked_sample,
  covariates = c("age", "sex", "race", "education", "name_commonness")
)

# Validate: does the weighted linked sample recover the population
# proportion urban?
validation <- validate_estimate(
  linked_data = linked_sample,
  population_data = census_population,
  outcome = "urban",
  adj_weights = ipw_no_urban$weights
)

validation
```

Because `urban` was held out from the propensity model, any improvement
in the estimate provides genuine evidence that the weights are working.

## Step 4: Correct for false matches

False matches can inflate estimates of rare transitions and attenuate
regression coefficients. When a validation variable is available (e.g.,
middle initials not used in linkage), we can estimate the false match rate
and apply a correction.

```{r false-match}
# Estimate false match rate from middle initial disagreement
mi_disagree <- linked_sample$middle_initial != linked_sample$middle_initial_mortality
false_match_rate <- mean(mi_disagree)
cat("Estimated false match rate:", round(false_match_rate, 3), "\n")
```

For a binary outcome, the adjustment factor from Equation 7 of the paper
corrects the observed rate:

```{r adjustment-factor}
# Suppose we're estimating the rate of a rare transition
# Among records with disagreeing middle initials (likely false matches),
# the transition rate reflects random pairing
observed_rate <- 0.08
rate_among_false <- 0.15   # estimated from validation subsample

af <- compute_adjustment_factor(
  observed_rate = observed_rate,
  false_match_rate = false_match_rate,
  rate_among_false = rate_among_false
)

cat("Adjustment factor:", round(af, 3), "\n")
cat("Corrected rate:", round(observed_rate * af, 4), "\n")
```

## Summary

The recommended workflow for linked data analysis:

1. **Compare** linked vs. population characteristics with `compare_linked()`
2. **Adjust** for missed matches using `weight_ipw()`, `weight_poststratify()`,
   or `weight_rake()`
3. **Validate** adjustments using `validate_estimate()` with a variable
   held out from the weighting model
4. **Correct** for false matches using `compute_adjustment_factor()` when a
   validation variable is available

## References

Breen, C.F. and Joo, W. (2026). "Population Research With Linked Data:
Methods for Inference." Working paper.
