---
title: "Introduction to linkinfer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to linkinfer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

`linkinfer` provides tools for researchers working with linked administrative
or census data. When records are linked across datasets, two types of errors
can affect inference:

1. **Missed matches** – records that should have linked but didn't, creating
   non‑random sample selection.
2. **False matches** – records incorrectly linked to the wrong individual,
   introducing measurement error.

This package helps diagnose and correct for these errors following the
framework of Breen and Joo (2026), "Population Research With Linked Data:
Methods for Inference."

A useful way to think about the workflow is in three steps. Researchers first
diagnose whether the linked sample is representative of the source population
and, if not, apply weighting methods to address missed matches. They then
evaluate whether false matches are likely to bias the estimate of interest
and, when possible, use an auxiliary validation variable to estimate and
correct that bias. Finally, they report both the original and corrected
estimates so that the implications of linkage error are transparent.

## Setup

```{r setup, include=FALSE}
# Reload the development version of linkinfer, ensuring no older version is attached
if ("linkinfer" %in% loadedNamespaces()) {
  detach("package:linkinfer", unload=TRUE, character.only=TRUE)
}
devtools::load_all('..')
```


```{r seed, echo=TRUE}
# Ensure reproducibility for the vignette
set.seed(1234)
```

## Example data

The package includes simulated data representing a census‑to‑linked records.
`population` is the full simulated census population, and `linked_sample`
is the subset that was successfully linked. In this example, the linked sample
is affected by both missed and false matches.

The simulation also includes a validation variable,
`middle_initial_match`, which is observed only for a subset of linked records.
This reflects a common empirical setting. Researchers often do not observe
ground truth for whether each linked pair is correct, but they may observe
an auxiliary variable that provides information about match quality for part
of the sample. Here, matching middle initials serve as a proxy for
higher-confidence links and are used to estimate how false matches may be
biasing the full-sample estimates.

```{r data}
# Simulate data (the function adds stochastic noise to `linked_prob`)
sim <- linked_data(n = 10000, p = 0.2, v1 = 0.5, v2 = 0.8)
population <- sim$population
linked_sample <- sim$linked

cat("Population size:", nrow(population), "\n")
cat("Linked sample size:", nrow(linked_sample), "\n")
cat("Match rate:", round(nrow(linked_sample) / nrow(population), 3), "\n")
```

## Step 1: Diagnose representativeness

The first step is to ask whether the linked sample resembles the population
from which it was drawn. If linkage success varies by age, sex, race,
education, or other characteristics, then inference based on the linked
sample may be biased even before considering false matches.

```{r compare}
comparison <- compare_linked(
  population = population,
  linked = linked_sample,
  vars = c("age", "sex", "race", "education", "urban", "name_commonness")
)
comparison
```

## Step 2: Adjust for missed matches

One way to correct for missed matches is to estimate each record’s probability of
appearing in the linked sample and then weight linked observations by the
inverse of that probability. This gives more influence to records from groups
that were harder to link.

### Inverse‑probability weighting

```{r ipw}
ipw_result <- weight_ipw(
  population = population,
  linked_indicator = "linked",
  covariates = c("age", "sex", "race", "education", "urban", "name_commonness"),
  method = "logistic"
)
ipw_result
```

### Diagnostics (balance plot)

After estimating weights, it is useful to check whether the weighted linked
sample is more similar to the population.

```{r ipw-diag}
diag <- weight_diagnostics(
  weights = ipw_result$weights,
  population = population,
  linked = linked_sample,
  covariates = ipw_result$covariates
)
diag
```

### Post‑stratification

If population totals are known for discrete subgroups, post-stratification
offers a simpler correction.

```{r poststrat}
ps_result <- weight_poststratify(
  population = population,
  linked = linked_sample,
  strata_vars = c("sex", "race")
)
ps_result

ps_diag <- weight_diagnostics(
  weights = ps_result$weights,
  population = population,
  linked = linked_sample,
  covariates = c("sex", "race")
)
ps_diag
```

### Raking

When exact joint cells become sparse, raking provides a more flexible way to
align the linked sample to known marginal distributions.

```{r raking}
rake_result <- weight_rake(
  linked_data = linked_sample,
  population = population,
  vars = c("sex", "race", "education")
)
rake_result
```

## Step 3: Adjust for false matches using a validation variable

False matches can bias estimates because some linked records are assigned to
the wrong individual. The adjustment-factor approach uses a validation
variable, not used in linkage itself, to estimate the extent of this bias in
a validation subsample and then applies that correction to the full linked
sample.

Let the observed rate in the linked sample be

\[
R' = R_{\text{true}}(1 - f_r) + R_{\text{false}} f_r,
\]

where \(R_{\text{true}}\) is the rate among true matches,
\(R_{\text{false}}\) is the rate among false matches, and \(f_r\) is the
false-match rate. Rearranging gives

\[
R_{\text{true}} = \frac{R' - R_{\text{false}} f_r}{1 - f_r}.
\]

Because these quantities are not observed in the full linked sample,
`linkinfer` estimates a correction in a validation subsample. In this
vignette, `middle_initial_match` plays that role. Since middle initial was
not used in matching, agreement in middle initial provides information about
whether a link is likely to be true.

Within the validation subsample, the adjustment factor is defined as

\[
\widehat{\mathrm{AF}} = \frac{\hat{R}_{\text{true}}}{\hat{R}},
\]

where \(\hat{R}_{\text{true}}\) is the estimate based on the
higher-confidence links and \(\hat{R}\) is the estimate based on all
validation links. The corrected full-sample estimate is then

\[
\hat{R}_{\text{adj}} = \hat{R}_{\text{full}} \times \widehat{\mathrm{AF}}.
\]

This method keeps the full linked sample for estimation while using the
validation subset to calibrate the likely bias from false matches. Its key
assumption is that the proportional bias observed in the validation subsample
is informative about the proportional bias in the full linked sample.

The function `adjust_estimate()` implements the adjustment-factor correction
for both descriptive outcomes and model-based estimates. By default,
model-based adjustments use ordinary least squares via `lm()`, but users may
supply any fitting function that accepts a `formula`, a `data` frame, and a
`weights` argument, such as `glm()`, `coxph()`, or `lmer()`. This is done
through the `model_fit` argument, together with a custom
`extract_estimate` function that returns the scalar quantity of interest from
the fitted model.

### Descriptive adjustment

We first illustrate the adjustment for a descriptive outcome. In this example,
the outcome is `race_diff`, and the function uses the validation variable
`middle_initial_match` to estimate an adjustment factor and apply it to the
weighted full-sample estimate.

```{r adjust-race-diff}
adjust_res <- adjust_estimate(
  weight_result = ipw_result,
  linked_data = linked_sample,
  population_data = population,
  outcome = "race_diff",
  validation_var = "middle_initial_match",
  n_boot = 200
)
adjust_res
```

### Regression‑based adjustment (OLS coefficient for Black vs White)

We next illustrate the same logic for a regression coefficient. Here the
quantity of interest is the coefficient on race_fBlack from an OLS
regression of age_at_death on race. The adjustment factor is estimated from
the validation subsample and then applied to the corresponding coefficient in
the full linked sample.

```{r adjust-regression, warning=FALSE, message=FALSE}
# Create a factor for race with White as reference
linked_sample$race_f <- factor(linked_sample$race, levels = c("White", "Black", "Other"))

model_form <- age_at_death ~ race_f

adjust_reg <- adjust_estimate(
  weight_result = ipw_result,
  linked_data = linked_sample,
  population_data = population,
  outcome = NULL,
  validation_var = "middle_initial_match",
  model_formula = model_form,
  extract_estimate = function(fit) coef(fit)["race_fBlack"],
  n_boot = 200
)
adjust_reg
```

## Summary

1. **Compare** linked vs. population characteristics with `compare_linked()`.
2. **Adjust** for missed matches using `weight_ipw()`, `weight_poststratify()`, or `weight_rake()`.
3. **Correct** for false matches using `adjust_estimate()` (validation‑variable based adjustment).

## References

Breen, C.F. and Joo, W. (2026). "Population Research With Linked Data:
Methods for Inference." Working paper.
